---
title: "Modeling Multilevel Data"
subtitle: "A Comparative Analysis of Hierarchical and Panel Econometric Approaches with Bayesian and Frequentist Perspectives in Financial Applications"
author: "Barry Quinn"
editor: visual
---

This chapter delves into the intricacies of modeling multilevel data, specifically focusing on comparing hierarchical models and panel econometrics within both Bayesian and frequentist perspectives. We will discuss each approach's underlying assumptions, strengths, weaknesses, and their application in finance. By understanding these methods, researchers and practitioners alike can make more informed decisions when selecting appropriate techniques for analyzing complex financial datasets.

# Introduction

Financial research often involves handling multilevel data structures where observations are nested or cross-classified across different levels (e.g., firms, industries, time periods). The choice of an adequate statistical methodology is crucial as it impacts estimation accuracy and subsequent policy recommendations. This chapter explores two prominent approaches -- hierarchical modeling and panel econometrics -- along with Bayesian and frequentist viewpoints.

## Group-Level Data Models

Before diving into specific modeling techniques, we introduce basic concepts related to group-level data. Let yij denote the observation i from level j, where i = 1, ..., ni and j = 1, ..., J. In this setup, traditional single-level regression models assume that all observations come from one homogeneous population. However, this assumption may not hold true in many cases, leading us to consider alternative frameworks capable of accounting for heterogeneity across groups.

## Hierarchical Linear Models (HLMs)

Hierarchical linear models, also known as mixed effects models or multi-level models, explicitly account for the variation between higher-level units by incorporating random components at multiple levels. HLMs have several advantages over classical regression techniques:

-   They allow for explicit modeling of correlations among lower-level units within groups;
-   They enable estimates of variance components at various levels;
-   They provide shrinkage estimators which borrow strength from similar groups, improving prediction performance;
-   They accommodate unbalanced designs with varying numbers of lower-level units per group.

## Panel Econometric Methods

Panel econometrics focuses on longitudinal data analysis using fixed effect and random effect models. These methods address potential endogeneity issues arising due to omitted variable bias and dynamic panels. Key features include:

-   Accounting for individual-specific unobserved heterogeneity through either fixed or random effects;
-   Capturing serial correlation via lagged dependent variables or error terms;
-   Allowing for robust inference using clustered standard errors.

## Bayesian vs Frequentist Perspective

From a philosophical standpoint, Bayesian and frequentist statisticians differ in their interpretation of probability. While frequentists view probability as long-run relative frequency, Bayesians interpret it as degree of belief based on prior knowledge. Consequently, these contrasting views lead to disparities in how parameters are estimated and interpreted.

## Financial Applications

Multilevel modeling has found extensive use in financial research, particularly in corporate finance, asset pricing, and risk management. Some examples include:

-   Analyzing firm-level data to estimate industry-specific cost functions while controlling for firm characteristics;
-   Examining portfolio returns to assess market efficiency after adjusting for cross-sectional dependence;
-   Investigating credit rating agencies' consistency in assigning ratings across countries and time periods.

## Discussion and Future Directions

While both hierarchical modeling and panel econometrics offer valuable insights into multilevel data analysis, they exhibit distinct characteristics requiring careful consideration before implementation. Moreover, recent advancements in computational algorithms facilitate seamless integration of Bayesian and frequentist paradigms, opening up new possibilities for sophisticated financial modeling. As such, further exploration of these topics promises exciting developments in our quest to better understand complex financial phenomena.

## Critiquing the models

### Fixed Effect Models

Fixed effect models (FEM) are widely used for addressing unobserved heterogeneity at the unit level. They capture nuisance parameters by including dummy variables for each entity. Advantages of FEM include:

-   Consistent parameter estimates even if the unobserved effects are correlated with covariates;
-   Within-group comparisons allowing for causal inferences under certain conditions;
-   Robustness to misspecification of functional forms compared to pooled OLS.

However, there are limitations associated with FEM:

-   Loss of degrees of freedom due to inclusion of numerous indicator variables, potentially affecting precision;
-   Assumption of zero mean for unobserved effects might be violated, resulting in biased coefficient estimates;
-   Cannot directly incorporate time-invariant predictors without resorting to first difference transformation or other approximations.

### Random Effect Models

Random effect models (REM), alternatively referred to as mixed effects models, treat unobserved heterogeneity as stochastic rather than deterministic entities. REM provides benefits such as:

-   Efficient estimation since only deviations from the overall mean need to be estimated for each unit;
-   Ability to model temporal dynamics using random slope coefficients;
-   Preservation of degrees of freedom compared to FEM.

Conversely, drawbacks of REM encompass:

-   Strict exogeneity assumption required for consistent estimation;
-   Sensitivity to specification of distributional form for random effects;
-   Potential inconsistency if unobserved effects are correlated with covariates (violation of "random effects" assumption).

### Hierarchical Models

Hierarchical linear models (HLMs), as previously discussed, allow for multiple levels of nesting and cross-classification. Compared to FEM and REM, HLM offers unique advantages:

-   Capability to handle complex data structures involving three or more levels;
-   Direct estimation of variance components at each level;
-   Shrinkage estimators providing improved predictions, especially for small clusters;
-   Increased power in detecting significant effects across groups.

Nevertheless, HLM entails its own set of challenges:

-   Computationally intensive due to iterative maximum likelihood procedures or Markov Chain Monte Carlo simulations;
-   Prone to identification problems when high correlations exist between predictors at different levels;
-   Requires careful justification of assuming normality for residuals and random effects distributions.

## Verdict

In conclusion, no single model outperforms others universally. Researchers must carefully evaluate their dataset's structure, research questions, and theoretical background to determine whether FEM, REM, or HLM is most suitable for their purposes. Additionally, sensitivity analyses should be conducted to ensure results' robustness across different modeling choices.

## Excercise

Sure, let's generate simulated data representing annual sales growth rates for subsidiaries belonging to different parent companies operating in various sectors. Our goal is to compare the performance of fixed effect models (FEM), random effect models (REM), and hierarchical linear models (HLM) in estimating sector-specific intercepts while controlling for parent company effects.

First, load necessary libraries and set seed for reproducibility:

```{R}
library(lme4) # For REM & HLM
library(plm)   # For FEM
set.seed(123)
```

Next, generate synthetic data consisting of sales growth rate `y`, parent company identifier `parent_id`, and sector classification `sector`. Assume there are 100 parents and five sectors, and each parent owns four subsidiaries observed annually over six years.

```{R}
# Number of parents, number of subsidiaries per parent, number of years, and number of sectors
n_parents <- 100  
subsidiary_per_parent <- 4
years <- 10
sectors <- 5

# Generate IDs
parent_id <- rep(1:n_parents, each=subsidiary_per_parent*years)  

# Generate time variable 
year <- rep(2015:2024, times=n_parents*subsidiary_per_parent)

# Generate sectors
sector <- sample(rep(1:sectors, each=subsidiary_per_parent*years/sectors))

# True sector effects  
beta <- rnorm(sectors, 0, 1)  

# Parent effects
alpha <- rnorm(n_parents, 0, 1) 

# Simulate growth rates   
sigma <- 0.05
y <- alpha[parent_id] + beta[sector] + rnorm(length(parent_id), sd=sigma)

# Create dataset
data <- data.frame(y, parent_id, sector, year)
```

Now fit the three models using respective packages:

```{R}
# Fit fixed effect model
# Create subsidiary id
sub_id <- 1:n_parents*subsidiary_per_parent*years
data$sub_id <- sub_id
index <- c("parent_id", "sub_id", "year")
# Model
fe_model <- plm(y ~ factor(sector), data, index = index)

# Fit random effect model
re_model <- lmer(y ~ factor(sector) + (1 | parent_id), data)

# Fit hierarchical linear model
hlm_model <- lmer(y ~ factor(sector) + (1|parent_id/sector), data)
```

Compare the estimated coefficients:

```{R}
cat("Fixed Effect Estimates:\n")
print(coef(fe_model)[-1], digits = 3)
cat("\n\nRandom Effect Estimates:\n")
print(fixef(re_model), digits = 3)
cat("\n\nHierarchical Linear Model Estimates:\n")
print(ranef(hlm_model)$parent_id[[1]], digits = 3)
print(ranef(hlm_model)$parent_id$sector, digits = 3)
```

You should observe that although all three models produce comparable estimates, the standard errors vary due to their differing assumptions about the nature of unobserved heterogeneity. Perform additional diagnostic checks and choose the best fitting model according to your problem requirements.

Keep in mind that this example represents a simplified scenario with only one explanatory variable. Real-world situations typically involve multiple predictors and require rigorous preprocessing steps, hypothesis testing, and validation procedures. Nevertheless, this exercise serves as a starting point towards grasping fundamental distinctions amongst FEM, REM, and HLM.

## Another excersise

Certainly! Let's adapt the exercise to a corporate finance context and implement it using R with a Bayesian approach. We'll simulate data reflecting a scenario often encountered in corporate finance, such as the effect of investment in research and development (R&D) on the financial performance of companies across different industries.

### Scenario:

-   **Dependent Variable**: Financial performance of companies (e.g., return on assets).
-   **Independent Variable**: Investment in R&D.
-   **Groups**: Different industries.

We will fit three models using R's `brms` package, which allows for Bayesian modeling: 1. **Fixed Effects Model**: Treat industry effects as fixed. 2. **Random Effects Model**: Treat industry effects as random. 3. **Multilevel Model**: Industry-specific random intercepts and possibly random slopes for R&D investment.

### R Code Implementation:

First, install and load the necessary package:

```{r}
#install.packages("brms")
library(brms)
```

### Step 1: Data Simulation

```{r}
set.seed(42)
n_companies <- 300
n_industries <- 10

# Simulating companies across different industries
industries <- factor(sample(1:n_industries, n_companies, replace = TRUE))

# Simulating R&D investment (independent variable)
rd_investment <- rnorm(n_companies, mean = 100, sd = 20)

# Simulating company performance (dependent variable)
# Assume a base performance, a positive effect of R&D, and some noise
industry_effect <- rnorm(n_industries, mean = 0, sd = 5) # random industry effect
performance <- 50 + 0.5 * rd_investment + industry_effect[as.numeric(industries)] + rnorm(n_companies, mean = 0, sd = 10)

data <- data.frame(Industry = industries, RDInvestment = rd_investment, Performance = performance)
```

### Step 2: Model Fitting

```{r}
# Bayesian Fixed Effects Model
model_fe <- brm(Performance ~ RDInvestment + factor(Industry), data = data, family = gaussian(), prior = c(set_prior("normal(0,10)", class = "b")))

# Bayesian Random Effects Model
model_re <- brm(Performance ~ RDInvestment + (1 | Industry), data = data, family = gaussian(), prior = c(set_prior("normal(0,10)", class = "b")))

# Bayesian Multilevel Model
model_ml <- brm(Performance ~ RDInvestment + (1 + RDInvestment | Industry), data = data, family = gaussian(), prior = c(set_prior("normal(0,10)", class = "b")))
```

### Step 3: Model Comparison and Interpretation

```{r}
summary(model_fe)
summary(model_re)
summary(model_ml)
```

### Execution and Analysis

-   Execute the R code in an environment with `brms` installed.
-   Examine the summaries of each model.
-   Focus on the coefficients for R&D investment and how they vary across models, especially looking at the industry-specific effects in the multilevel model.

This exercise will illustrate how Bayesian models can be applied in a corporate finance context, emphasizing the differences between fixed effects, random effects, and multilevel modeling approaches.
